Quantifying the importance of an inference model in Bayesian phylogenetics

### 2020-07-28 pirouette_article.tex
Process feedback. Thanks @rsetienne. Fix #111

\begin{figure}[H]
  \includegraphics[width=\textwidth]{pirouette_example_30/errors.png}
  \caption{
    \new{The impact of the tree prior for the example tree in Figure \ref{fig:dd_tree}. The alignment for this true tree was generated using a JC69 substitution model and strict clock model. For inferring the tree from this alignment in the 'generative' scenario the same substitution and clock models were used, and a Yule tree prior (this is the assumed generative model because the real generative model is of course unknown). In the 'best' scenario the best-fitting candidate models (JC69 substitution model, relaxed log-normal clock model and Yule tree prior) were used. For the twin tree, the same models were chosen, except that BD was used to generate and infer the twin tree.
    The twin distributions show the baseline inference error.}
    Vertical dashed lines show the median error value per distribution.
  }
  \label{fig:example_30}
\end{figure}

>>>>

\begin{figure}[H]
  \includegraphics[width=0.9\textwidth]{pirouette_example_30/errors.png}
  \caption{
    \new{
      The impact of the tree prior for 
      the example tree in Figure \ref{fig:dd_tree}. 
      The alignment for this true tree was generated using a JC substitution model 
      and strict clock model. 
      For inferring the tree from this alignment in the 'generative' scenario 
      the same substitution and clock models were used, 
      and a Yule tree prior (this is the assumed generative model,
      because the real generative model is of course unknown). 
      In the 'best' scenario the best-fitting candidate 
      models (TN substitution model, strict clock model and Yule tree prior) 
      were used. For the twin tree, the same models were chosen, 
      except that BD was used to generate and infer the twin tree.
      The twin distributions show the baseline inference error.
    }
    Vertical dashed lines show the median error value per distribution.
  }
  \label{fig:example_30}
\end{figure}

### 2020-07-28 pirouette_supplement.tex
Process feedback. Thanks @rsetienne. Fix #111

From this 'true tree', a 'true alignment' is simulated, using
the JC69 nucleotide substitution model and a strict clock model.
The resulting alignment is shown at the center-left
of figure \ref{fig:pipeline}.

From the 'true alignment' the generative inference model is run.
Of course, it cannot be the actual (DD) model. Instead, the
default BEAST2 inference model is used, which assumes a JC69 nucleotide
substitution model, a strict clock model and a Yule tree model.
The resulting posterior trees are shown in the
center-left panel of figure \ref{fig:pipeline}.

@@

Based on the 'true alignment' (center-left panel), the candidate model
with the highest marginal likelihood is determined, from a set of 
15 models. The set of models consists of all combinations
of all 4 nucleotides substitution models (JC69, HKY, TN93, GTR),
all 2 clock models (strict and relaxed log-normal) and 2 birth-death
models (Yule and Birth-Death), except the inference model used as the
generative model (JC69, strict clock, Yule). The inference model that had the
highest evidence (as shown in Table \ref{tab:evidence_true}) was 
the inference model with a JC69 nucleotide substitution model,
a relaxed log-normal clock model and a Yule tree model.
The resulting posterior trees are shown in the second panel
of the third row of posteriors in figure \ref{fig:pipeline}.

@@

From this 'twin tree', a 'twin alignment' was simulated, using
the JC69 nucleotide substitution model and a strict clock model.
The resulting alignment is shown in the center-right panel
of figure \ref{fig:pipeline}.

From the 'twin alignment' the generative inference model is run as well.
Also here, the default BEAST2 inference model is used, 
which assumes a JC69 nucleotide
substitution model, a strict clock model and a Yule tree model.
The resulting posterior trees are shown in the
third panel of the third row of figure \ref{fig:pipeline}.
From this 'generative twin' posterior, 
the difference between each of its trees is
compared to the 'twin tree' (top-right panel), using the nLTT statistic,
resulting in the error distribution shown in the third panel
of the bottom row of figure \ref{fig:pipeline}.

Based on the 'twin alignment' (center-right panel), the candidate model
with the highest marginal likelihood is determined, from the same 
set of 15 candidate models.The inference model that had the
highest evidence (as shown in Table \ref{tab:evidence_twin}) was 
the inference model with a JC69 nucleotide substitution model,
a strict clock model and a BD tree model (note: this indeed matches how
the twin tree and twin tree were simulated).
From the 'twin alignment' this best candidate inference model is run.
The resulting posterior trees are shown in the fourth panel
of the third row of posteriors in figure \ref{fig:pipeline}.

@@

The ESSes of the 'true' pipeline for the generative model
are shown in Table \ref{tab:esses_gen}.
From the estimated parameters, one can deduce that
the JC69 nucleotide substitution model was used (no
estimated parameter needed), a strict clock model was 
used (again, no parameter needed to be estimated)
and a Yule tree prior is used ('Yule model' and 'birthRate' are estimated).
Note that although the actual true tree is created by a DD process, 
the default and standard Yule tree model is used as the closest
standard tree model.

@@

The ESSes of the 'true' pipeline for the best candidate model
are shown in Table \ref{tab:esses_best}. 
From the names of the estimated parameters, it is clear that
the best candidate model has had a TN93 nucleotide substitution 
model (due to 'kappa1', 'kappa2', 'freqParameter.1', 
'freqParameter.2', 'freqParameter.3', 'freqParameter.4')
a strict clock (no parameter needed to be estimated) and a BD 
model ('BirthDeath', 'BDBirthRate' and 'BDDeathRate').

@@

From the names of the estimated parameters, it is clear that
the best candidate model for the twin tree
is JC69 nucleotide substitution 
model (no parameter needed to be estimated),
a strict clock (no parameter needed to be estimated) and a BD 
model ('BirthDate', 'BDBirthRate' and 'BDDeathRate').
Note that this matches the actual process of how the twin tree
and twin alignment are generated.

@@

The marginal likelihood (or evidence) data for the model comparison
performed in the 'true' pipeline is shown in Table \ref{tab:evidence_true}.
The best (that is, the one with the highest model weight)
candidate model assumes a TN93 nucleotide substitution model,
a strict clock and a BD model.

@@

The marginal likelihood (or evidence) data for model comparison
performed in the 'twin' pipeline is shown in Table \ref{tab:evidence_twin}.
The best (that is, the one with the highest model weight)
candidate model assumes a JC69 nucleotide substitution model,
a strict clock and a BD model.
Note that this matches the actual process of how the twin tree
and twin alignment are generated.

>>>>

From this 'true tree', a 'true alignment' is simulated, using
the JC nucleotide substitution model and a strict clock model.
The resulting alignment is shown at the center-left
of figure \ref{fig:pipeline}.

From the 'true alignment' the generative inference model is run.
Of course, it cannot be the actual (DD) model. Instead, the
default BEAST2 inference model is used, which assumes a JC nucleotide
substitution model, a strict clock model and a Yule tree model.
The resulting posterior trees are shown in the
center-left panel of figure \ref{fig:pipeline}.

@@

Based on the 'true alignment' (center-left panel), the candidate model
with the highest marginal likelihood is determined, from a set of 
15 models. The set of models consists of all combinations
of all 4 nucleotides substitution models (JC, HKY, TN, GTR),
all 2 clock models (strict and relaxed log-normal) and 2 birth-death
models (Yule and Birth-Death), except the inference model used as the
generative model (JC, strict clock, Yule). The inference model that had the
highest evidence (as shown in Table \ref{tab:evidence_true}) was 
the inference model with a JC nucleotide substitution model,
a relaxed log-normal clock model and a Yule tree model.
The resulting posterior trees are shown in the second panel
of the third row of posteriors in figure \ref{fig:pipeline}.

@@

From this 'twin tree', a 'twin alignment' was simulated, using
the JC nucleotide substitution model and a strict clock model.
The resulting alignment is shown in the center-right panel
of figure \ref{fig:pipeline}.

From the 'twin alignment' the generative inference model is run as well.
Also here, the default BEAST2 inference model is used, 
which assumes a JC nucleotide
substitution model, a strict clock model and a Yule tree model.
The resulting posterior trees are shown in the
third panel of the third row of figure \ref{fig:pipeline}.
From this 'generative twin' posterior, 
the difference between each of its trees is
compared to the 'twin tree' (top-right panel), using the nLTT statistic,
resulting in the error distribution shown in the third panel
of the bottom row of figure \ref{fig:pipeline}.

Based on the 'twin alignment' (center-right panel), the candidate model
with the highest marginal likelihood is determined, from the same 
set of 15 candidate models.The inference model that had the
highest evidence (as shown in Table \ref{tab:evidence_twin}) was 
the inference model with a JC nucleotide substitution model,
a strict clock model and a BD tree model (note: this indeed matches how
the twin tree and twin tree were simulated).
From the 'twin alignment' this best candidate inference model is run.
The resulting posterior trees are shown in the fourth panel
of the third row of posteriors in figure \ref{fig:pipeline}.

@@

The ESSes of the 'true' pipeline for the generative model
are shown in Table \ref{tab:esses_gen}.
From the estimated parameters, one can deduce that
the JC nucleotide substitution model was used (no
estimated parameter needed), a strict clock model was 
used (again, no parameter needed to be estimated)
and a Yule tree prior is used ('Yule model' and 'birthRate' are estimated).
Note that although the actual true tree is created by a DD process, 
the default and standard Yule tree model is used as the closest
standard tree model.

@@

The ESSes of the 'true' pipeline for the best candidate model
are shown in Table \ref{tab:esses_best}. 
From the names of the estimated parameters, it is clear that
the best candidate model has had a TN nucleotide substitution 
model (due to 'kappa1', 'kappa2', 'freqParameter.1', 
'freqParameter.2', 'freqParameter.3', 'freqParameter.4')
a strict clock (no parameter needed to be estimated) and a BD 
model ('BirthDeath', 'BDBirthRate' and 'BDDeathRate').

@@

From the names of the estimated parameters, it is clear that
the best candidate model for the twin tree
is JC nucleotide substitution 
model (no parameter needed to be estimated),
a strict clock (no parameter needed to be estimated) and a BD 
model ('BirthDate', 'BDBirthRate' and 'BDDeathRate').
Note that this matches the actual process of how the twin tree
and twin alignment are generated.

@@

The marginal likelihood (or evidence) data for the model comparison
performed in the 'true' pipeline is shown in Table \ref{tab:evidence_true}.
The best (that is, the one with the highest model weight)
candidate model assumes a TN nucleotide substitution model,
a strict clock and a BD model.

@@

The marginal likelihood (or evidence) data for model comparison
performed in the 'twin' pipeline is shown in Table \ref{tab:evidence_twin}.
The best (that is, the one with the highest model weight)
candidate model assumes a JC nucleotide substitution model,
a strict clock and a BD model.
Note that this matches the actual process of how the twin tree
and twin alignment are generated.
